---
title: "MR stickleback"
author: "Jeff Walker"
date: "April 5, 2018"
output:
  bookdown::html_document2: default
  pdf_document: default
header-includes:
- \usepackage{tikz}
- \usetikzlibrary{positioning}
---

# History
```{r}
# reboot Aug 9 2013
# reboot II April 5 2018
```
The original motivation was a Mendelian Randomization re-analysis of
Marchinko, K.B., 2009. Predation's role in repeated phenotypic and genetic divergence of armor in threespine stickleback. Evolution, 63(1), pp.127-138.
The new goal is to add the Schluter xxx data.

We need a DAG model of the potential paths. The LaTeX script only creates an image in the pdf. So the script is run to create the pdf, then re-run to create the html/pdf.

<!--   \begin{tikzpicture} -->
<!-- 		\node (1) {$Z_{eda}$}; -->
<!-- 		\node[right=of 1] (2) {$X_{armor}$}; -->
<!-- 		\node[right=of 2] (3) {$Y_w$}; -->
<!-- 		\node[below=of 1] (4) {$U_g$}; -->
<!-- 		\node[below=of 2] (5) {$U_x$}; -->
<!-- 		\node[above=of 2] (6) {$X_{SL}$}; -->
<!-- 		\draw [->] (1) -- node[above,font=\footnotesize]{$\beta_1$} (2); -->
<!-- 		\draw [->] (2) -- node[above,font=\footnotesize]{$\beta_2$} (3); -->
<!-- 		\draw [->] (1) -- node[above,pos=.3,font=\footnotesize]{$\beta_3$} (5); -->
<!-- 		\draw [->] (5) -- node[below,font=\footnotesize]{$\beta_4$} (3); -->
<!-- 		\draw [->] (4) -- node[below,pos=.3,font=\footnotesize]{$\beta_5$} (2); -->
<!-- 		\draw [->] (4) -- node[below,font=\footnotesize]{$\beta_6$} (5); -->
<!-- 		\draw [->] (6) -- node[above,font=\footnotesize]{$\beta_7$} (3); -->
<!-- 		\draw [->] (1) -- node[above,font=\footnotesize]{$\beta_8$} (6); -->
<!-- 		\draw [->] (1) to[bend right=25] node[left,font=\footnotesize]{$r1$} (4); -->
<!-- 		\draw [->] (4) [bend left=25] to (1); -->
<!-- 		\draw [->] (2) to[bend left=25] node[right,font=\footnotesize]{$r2.z$} (6); -->
<!-- 		\draw [->] (6) [bend right=25] to (2); -->
<!-- 	\end{tikzpicture} -->

```{r path, echo=FALSE, fig.cap="Full model."}
dir_path <- "images/" # for console
dir_path <- "../images/" # for knit
image_path <- paste(dir_path, "path.png", sep="")
knitr::include_graphics(image_path)
  
```

MR estimates $\beta_2$. An unbiased Mendelian randomization (MR) estimate of $\beta_2$ assumes

1. $\beta_3 \beta_4 = 0$
2. $r \beta_6 \beta_4 = 0$

The estimate of the "direct selection on" $X_{armor}$ ($\beta_2$), conditioning on $Z_{eda}$, is an unbiased estimate of $\beta_2$ if

1. $\beta_5 \beta_6 \beta_4 = 0$

# Load libraries

```{r libraries, message=FALSE, warning=FALSE}
library(readxl)
library(lme4)
library(ggplot2)
library(data.table)

source("../R/replicate.R") # for knit, drop the "../" for console
source("../R/mendelian_randomization.R") # for knit, drop the "../" for console

```

# Import Rennison data
```{r Rennison}
dir_path <- "data/" # for console
dir_path <- "../data/" # for knit
file_name <- "Rennison et al archive.csv"
file_path <- paste(dir_path, file_name, sep='')
rennison <- data.table(fread(file_path))
rennison[, month:=factor(month, c('06-Sep', '06-Oct', '06-Nov'))]
rennison[, genotype:=factor(genotype)]

```

# Replicate Rennison
## Replicate size adjustment
### Using fixed effects model (ANCOVA)
It would be good to simulate process to know how sensitive result is to adjustment vs. simply using length as covariate in linear model.

The script attempts to use a function within data.table but limiting the analysis to length < 34.0. While the fit is only on data with length < 34.0, data.table is giving the predicted value for all data. This seems to work after re-starting.

```{r replicateRennisonSize}
# plot plates vs. size
qplot(x=length, y=raw.plate.total, data=rennison)
# check size adjustment - can't replicate
rennison[, adj.logplates_rep1:=ifelse(length < 34.0, size_corrected(raw.logplates[length < 34.0], length[length < 34.0], origin=34.0), raw.logplates), by=.(genotype)]
# limit adjusted plates to max of log(32)
rennison[, adj.logplates_rep1:=ifelse(adj.logplates_rep1 > log(32), log(32), adj.logplates_rep1)]

# check on the math of above - the above is not fitting on the subset of length < 34
  # x <- rennison[genotype=='CL' & length< 34, raw.logplates]
  # sl <- rennison[genotype=='CL' & length< 34, length]
# fit <- lm(x~sl)
# x_34 <- predict(fit, newdata=data.frame(sl=34.0))
# logplates.s <- x_34 + residuals(fit) 


```

The ANCOVA model doesn't replicate.
```{r replicateRennisonSizeANCOVA}

  fit <- lm(raw.logplates ~ length*genotype, data=rennison[length < 34.0,])

  # # residuals for all data (not just length <= 34)
  # rennison[, logplates.resid:=raw.logplates - predict(fit, data.frame(length=length, genotype=genotype))]
  # # predicted value for all data at length=34.0 mm
  # rennison[, logplat.hat34:=predict(fit, data.frame(length=34.0, genotype=genotype))]
  # # adjusted values
  # rennison[, logplates.s2b:=ifelse(length <= 34, logplat.hat34 + logplates.resid, raw.logplates)]
  
  # if length < 34.0 mm then value is predicted value at length=34 (conditional on genotype) plus residual value (conditional on length*genotype)
  rennison[, adj.logplates_rep2:=ifelse(length < 34,
                    predict(fit, data.frame(length=34.0, genotype=genotype)) + 
                    raw.logplates - predict(fit, data.frame(length=length, genotype=genotype)),
                    raw.logplates)]
  # limit adjusted plates to max of log(32)
  rennison[, adj.logplates_rep2:=ifelse(adj.logplates_rep2 > log(32), log(32), adj.logplates_rep2)]

```

### Replicate size adjustment using Rennison et al. script
E-mail from Rennison on April 08, 2018 states that online dataset not full dataset. Using Rennison script, I can compute the estimate of the breakpoint $k$ in the nls function and then re-run ANCOVA without the NLS. Same to 2nd to 3rd decimal place.

```{r rennison_code}
# change logplates to raw.logplates

fit <- nls(raw.logplates ~  (a + b * length) * as.numeric(genotype == "CC" & length<=k) + 
             (a + b * k) * as.numeric(genotype == "CC" & length>k) +
             (c + d * length) * as.numeric(genotype == "CL" & length<=k) +
             (c + d * k) * as.numeric(genotype == "CL" & length>k) +
             (e + f * length) * as.numeric(genotype == "LL" & length<=k) +
             (e + f * k) * as.numeric(genotype == "LL" & length>k),
           data=rennison, algorithm = "default",
           start=list(a=1, c=1, e=1, b=0.1, d=0.1, f=0.1, k=34), control=list(warnOnly=TRUE))

rennison[, adj.logplates_rep3:=ifelse(length < 34,
              predict(fit, data.frame(length=34.0, genotype=genotype)) + 
              raw.logplates - predict(fit, data.frame(length=length, genotype=genotype)),
              raw.logplates)]
  rennison[, adj.logplates_rep3:=ifelse(adj.logplates_rep3 > log(32), log(32), adj.logplates_rep3)]

# rerun my routine with k=34.639958.

  fit <- lm(raw.logplates ~ length*genotype, data=rennison[length < 34.639958,])
  rennison[, adj.logplates_rep4:=ifelse(length < 34,
                    predict(fit, data.frame(length=34.0, genotype=genotype)) + 
                    raw.logplates - predict(fit, data.frame(length=length, genotype=genotype)),
                    raw.logplates)]
  # limit adjusted plates to max of log(32)
  rennison[, adj.logplates_rep4:=ifelse(adj.logplates_rep4 > log(32), log(32), adj.logplates_rep4)]

  # compare
  qplot(x=adj.logplates, y=adj.logplates_rep4, data=rennison)
  qplot(x=adj.logplates_rep3, y=adj.logplates_rep4, data=rennison)
  
  rennison[, adj.plates_rep4:=exp(adj.logplates_rep4)]

```


### Using multilevel model

```{r rennisonSizeMLM}
  fit <- lmer(raw.logplates ~ length+ genotype + (1|genotype), data=rennison)
  

```


## Replicate multivariate selection differentials

### Bootstrap not including size adjustment, using Rennison data
The bootstrap implemented below does not bootstrap the whole process including the length standardization step. Re-sampling should account for the variability in frequency of CC and CL genotypes (so this frequency is not fixed every iteration)

```{r replicateRennisonDifferentials1}

# get frequencies of genotype x month within each month
freqs <- cbind(
  rennison[genotype!='LL', .(Nj=.N), by=.(genotype, month)],
  N=rep(rennison[genotype!='LL', .(N=.N), by=.(month)][, N], each=2))
freqs[, f:=Nj/N] 

# following Rennison code, use frequencies from Barrett 2008, which is larger dataset
freq_sept <- c(0.47, 0.53) # CC=0, CL=1
freq_oct <- c(0.7, 0.3)
freq_nov <- c(0.54, 0.46)

niter <- 9999
ycols <- c('adj.plates', 'eda_dom')
Xa <- rennison[month==levels(month)[1] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
# mv_selection_differential <- function(Xa, Xb, freq_a=NULL, freq_b=NULL, bootstrap=FALSE, niter=4999){
res1 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_sept, freq_b=freq_oct, bootstrap=TRUE, niter=niter)
apply(res1, 2, mean)
table1.rep1 <- apply(res1,2, quantile, c(0.025, 0.5, 0.975))

Xa <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[3] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
res2 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_oct, freq_b=freq_nov, bootstrap=TRUE)
table2.rep1 <- apply(res2,2, quantile, c(0.025, 0.5, 0.975))

# 
```

### Bootstrap not including size adjustment, using rep4 data
How much does including the size adjustment in the bootstrap matter in the error? For this, I need the estimate using my adjustment routine, not within the bootstrap as the basis of comparison

The bootstrap implemented below **does not bootstrap the whole process** including the length standardization step
```{r replicateRennisonDifferentials2}
# following Rennison code, use frequencies from Barrett 2008, which is larger dataset
freq_sept <- c(0.47, 0.53) # CC=0, CL=1
freq_oct <- c(0.7, 0.3)
freq_nov <- c(0.54, 0.46)

ycols <- c('adj.plates_rep4', 'eda_dom')
niter <- 9999

Xa <- rennison[month==levels(month)[1] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
# mv_selection_differential <- function(Xa, Xb, freq_a=NULL, freq_b=NULL, bootstrap=FALSE, niter=4999){
res1 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_sept, freq_b=freq_oct, bootstrap=TRUE, niter=niter)
table1.rep4.mean <- apply(res1, 2, mean)
table1.rep4 <- apply(res1,2, quantile, c(0.025, 0.5, 0.975))

Xa <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[3] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
res2 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_oct, freq_b=freq_nov, bootstrap=TRUE, niter=niter)
table2.rep4.mean <- apply(res2, 2, mean)
table2.rep4 <- apply(res2,2, quantile, c(0.025, 0.5, 0.975))
# 
```

### Bootstrap including size adjustment, using rep4 data

```{r replicateRennisonDifferentials3}
# following Rennison code, use frequencies from Barrett 2008, which is larger dataset
freq_sept <- c(0.47, 0.53) # CC=0, CL=1
freq_oct <- c(0.7, 0.3)
freq_nov <- c(0.54, 0.46)

ycols <- c('length', 'raw.logplates', 'eda_dom')
niter <- 9999

Xa <- rennison[month==levels(month)[1] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
# mv_selection_differential <- function(Xa, Xb, freq_a=NULL, freq_b=NULL, bootstrap=FALSE, niter=4999){
res1 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_sept, freq_b=freq_oct, adjust=TRUE, bootstrap=TRUE, niter=niter)
table1.rep4b.mean <- apply(res1, 2, mean)
table1.rep4b <- apply(res1,2, quantile, c(0.025, 0.5, 0.975))

Xa <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[3] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
res2 <- mv_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_oct, freq_b=freq_nov, bootstrap=TRUE, niter=niter)
table2.rep4.mean <- apply(res2, 2, mean)
table2.rep4b <- apply(res2,2, quantile, c(0.025, 0.5, 0.975))

  # compare
  table1.rep1 # original data
  table1.rep4 # my size adjusted
  table1.rep4b # bootstrapped size adjusted

```

# Rennison Mendelian Randomization

note that can use published estimates or meta-analytic estimates of B_zx

## Using size-adjusted lateral plate counts following Rennison analysis

### Rennison adjustments old code
Doesn't use the 2008 Barrett frequencies of genotype
```{r rennisonMR}
inc <- c(which(rennison[eda_add!=-1, month]=='06-Sep'), which(rennison[eda_add!=-1, month]=='06-Oct'))
x <- rennison[inc, adj.plates]
y <- rennison[inc, month]
y <- factor(as.character(y), c('06-Sep', '06-Oct'))
z <- rennison[inc, eda_dom]
res <- mendelian_randomization(x,y,z,std=TRUE, bootstrap=TRUE)
table1.mr.rep1 <- apply(res,2, quantile, c(0.025, 0.5, 0.975))

inc <- c(which(rennison[eda_add!=-1, month]=='06-Oct'), which(rennison[eda_add!=-1, month]=='06-Nov'))
x <- rennison[inc, adj.plates]
y <- rennison[inc, month]
y <- factor(as.character(y), c('06-Oct', '06-Nov'))
z <- rennison[inc, eda_dom]
res <- mendelian_randomization(x,y,z,std=TRUE, bootstrap=TRUE)
table2.mr.rep1 <- apply(res,2, quantile, c(0.025, 0.5, 0.975))

```

### Rennison adjustment with modified multivariate selection differential code

```{r rennisonMRnew}
# following Rennison code, use frequencies from Barrett 2008, which is larger dataset
freq_sept <- c(0.47, 0.53) # CC=0, CL=1
freq_oct <- c(0.7, 0.3)
freq_nov <- c(0.54, 0.46)

ycols <- c('adj.plates', 'eda_dom')
Xa <- rennison[month==levels(month)[1] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
# mr_selection_differential <- function(Xa, Xb, codes=NULL, freq_a=NULL, freq_b=NULL, bootstrap=FALSE, niter=4999){
res1 <- mr_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_sept, freq_b=freq_oct, bootstrap=TRUE)
table1.mr.rep2 <- apply(res1,2, quantile, c(0.025, 0.5, 0.975))

Xa <- rennison[month==levels(month)[2] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xa, old='eda_dom', new='genotype')
Xb <- rennison[month==levels(month)[3] & eda_add!=-1, .SD, .SDcols=ycols]
setnames(Xb, old='eda_dom', new='genotype')
res2 <- mr_selection_differential(Xa, Xb, codes=c(0,1), freq_a=freq_oct, freq_b=freq_nov, bootstrap=TRUE)
table2.mr.rep2 <- apply(res2,2, quantile, c(0.025, 0.5, 0.975))

```

# Marchinko analysis
## Import Marchinko data

```{r marchinko}
dir_path <- "data/" # for console
#dir_path <- "../data/" # for knit
file_name <- "Marchinko_PaxtonData.xls"
file_path <- paste(dir_path, file_name, sep='')

marchinko <- data.table(read_excel(file_path))
	# columns are
		# population
		# family
		# treatment {no = -insect, pred = +insect}
		# individual
		# standlength
		# ant.dorspine
		# sec.dorspine
		# pelspine.len
		# pelgirdle.len
		# plate.number
		# eda.genotype {AA, Aa, aa}
		# pelgirdle.presence {1=yes, 0 = no}
		# pelspine.presence {1 = yes, 0 = no}

  marchinko[, family:=factor(family)]
  marchinko[, treatment:=factor(treatment, c('no','pred'))]
# additive model of eda
marchinko[, eda_add:=ifelse(eda.genotype=="aa",-1,ifelse(eda.genotype=="Aa",0,1))]
# dominance model of eda
marchinko[, eda_dom:=ifelse(eda.genotype=="aa",0,1)]

```

## Replicate Marchinko

### Replicate using fixed effects model

Replication requires scaling armor variables within families and replacing values=0.0 with NA. Still, I cannot replicate *ant.dorspine.s*
```{r replicate}
  table_1_replicate <- round(replicate_marchinko(marchinko),3)
```

### Re-analysis using multi-level model

```{r mlm}

y_list <- c('ant.dorspine', 'sec.dorspine', 'pelspine.len', 'pelgirdle.len')
table_1_mlm <- numeric(length(y_list))
for(j in 1:length(y_list)){
  y <- y_list[j]
  # fit to get treatment effect
  model_formula <- formula(paste(y, " ~ standlength + treatment + (standlength|family)", sep=''))
  fit <- lmer(model_formula, data=marchinko)
  b_xy <- coefficients(summary(fit))['treatmentpred', 'Estimate']
  # get sd from control group only
  model_formula <- formula(paste(y, " ~ standlength + (standlength|family)", sep=''))
  fit <- lmer(model_formula, data=marchinko[treatment=='no',])
  sd_x <- unlist(as.data.table(VarCorr(fit))[grp=='Residual','sdcor'])
  sd_x <- 1
  table_1_mlm[j] <- b_xy/sd_x
}
table_1_mlm

```

## Mendelian randomization of Marchinko armor data
### MR of lateral plate data

```{r marchinkoLateralPlate}
# Scale plate count
# first explore, row 192, 19, 52 are conspicuous outliers
qplot(x=standlength, y=plate.number, data=marchinko)
fit <- lm(plate.number ~ standlength, data=marchinko)
fitlog <- lm(plate.number ~ log(standlength), data=marchinko)
plot(fit)
plot(fitlog)
coefficients(lmer(plate.number ~ standlength + (1|family), data=marchinko[treatment=='no',]))


xcol <- 'plate.number'
zcol <- 'eda.genotype'
covcol <- 'standlength'
randcol <- 'family'

Xa <- marchinko[treatment=='no', .SD, .SDcols=c(xcol, zcol, covcol, randcol)]
Xb <- marchinko[treatment=='pred', .SD, .SDcols=c(xcol, zcol, covcol, randcol)]

#2SLS
# include standardlength in both?
qplot(x=eda_add, y=plate.number, data=marchinko)
fit <- lmer(plate.number ~ standlength + eda_add + (1|family), data=marchinko[treatment=='no',])
x_res <- predict(fit, newdata=marchinko) - marchinko$plate.number

fit <- lmer(x_res ~ treatment*standlength + (1|family), data=marchinko)
b_xy_mr <- coefficients(summary(fit))['treatmentpred', 'Estimate']

# get sd_x from control group only and from non-RM model
fit <- lmer(plate.number ~ standlength + (standlength|family), data=marchinko[treatment=='no',])
sd_x <- unlist(as.data.table(VarCorr(fit))[grp=='Residual','sdcor'])

b_xy_mr_prime <- b_xy_mr/sd_x

#2SLS
niter <- 2000
B_colnames <- c('b_zy', 'b_zx', 'r_zx', 'b_xy', 'b_xy_prime', 'beta_xy', 'beta_xy_prime')
B <- matrix(NA, nrow=niter, ncol=length(B_colnames))
colnames(B) <- B_colnames
inc <- 1:nrow(marchinko)
for(iter in 1:niter){
  # first use residuals of plates on length
  dt <- marchinko[inc,]
  fit <- lmer(plate.number ~ standlength*eda.genotype + (1|family), data=dt[treatment=='no',], na.action="na.exclude")
  dt[, plate_res:=predict(fit, newdata=dt)-plate.number]

  # selection differential on genotype
  fit <- lmer(eda_add ~ treatment + (1|family), data=dt, na.action="na.exclude")
  B[iter, 'b_zy'] <- coefficients(summary(fit))['treatmentpred', 'Estimate']

  # selection differential on phenotype
  fit <- lmer(plate_res ~ treatment + (1|family), data=dt, na.action="na.exclude")
  B[iter, 'b_xy'] <- coefficients(summary(fit))['treatmentpred', 'Estimate']

  # stage 1
  fit1 <- lmer(plate_res ~ eda_add + (1|family), data=dt[treatment=='no',], na.action="na.exclude")
  B[iter, 'b_zx'] <- coefficients(summary(fit1))['eda_add', 'Estimate']
  x_res <-predict(fit1, newdata=dt) - dt$plate_res
  
  fit2 <- lmer(x_res ~ treatment + (1|family), data=dt, na.action="na.exclude")
  B[iter, 'beta_xy']  <- coefficients(summary(fit2))['treatmentpred', 'Estimate']
  
  # get sd_x from control group only
  fit <- lmer(plate_res ~ 1 + (1|family), data=dt[treatment=='no',], na.action="na.exclude")
  sd_x <- unlist(as.data.table(VarCorr(fit))[grp=='Residual','sdcor'])
  
  B[iter, 'b_xy_prime']  <- B[iter, 'b_xy']/sd_x
  B[iter, 'beta_xy_prime']  <- B[iter, 'beta_xy']/sd_x
  inc <- sample(1:nrow(marchinko), replace=TRUE)
}

apply(B, 2, quantile, c(0.025, 0.5, 0.975), na.rm=TRUE)


```

## MR of armor complex
It doesn't make sense to compute this for each armor trait separately because then by definition there assumptions are violated (Z has a path to W independent of X). Possible solution here: "Multivariable Mendelian randomization: the use of pleiotropic genetic variants to estimate causal effects", doi: 10.1093/aje/kwu283.

$\beta_{zy}$ is the treatment effect on *eda* genotype. 
```{r betaZY}
# additive model of eda
marchinko[, eda_add:=ifelse(eda.genotype=="aa",-1,ifelse(eda.genotype=="Aa",0,1))]
# dominance model of eda
marchinko[, eda_dom:=ifelse(eda.genotype=="aa",0,1)]

fit <- lmer(eda_add ~ treatment + (1|family), data=marchinko)
b_zy <- coefficients(summary(fit))['treatmentpred', 'Estimate']

```

$\beta_{zx}$ is the effect of *eda* genotype on armor phenotype in the control group.

```{r betaZX}
p <- length(y_list)
b_zx <- numeric(p)
sd_x <- numeric(p)
b_zx_prime <- numeric(p)
table_1_mr <- numeric(p)
table_1_mr_prime <- numeric(p)
for(j in 1:length(y_list)){
  y <- y_list[j]
  # fit to get treatment effect
  model_formula <- formula(paste(y, " ~ standlength + eda_add + (standlength|family)", sep=''))
  fit <- lmer(model_formula, data=marchinko[treatment=='no'])
  b_zx[j] <- coefficients(summary(fit))['eda_add', 'Estimate']
  
  # get sd_x from control group only and from non-RM model
  model_formula <- formula(paste(y, " ~ standlength + (standlength|family)", sep=''))
  fit <- lmer(model_formula, data=marchinko[treatment=='no',])
  sd_x[j] <- unlist(as.data.table(VarCorr(fit))[grp=='Residual','sdcor'])
  
  table_1_mr[j] <- b_zy/b_zx[j]
  b_zx_prime[j] <- b_zx[j]*(sd_z/sd_x[j])
  table_1_mr_prime[j] <- b_zy_prime/b_zx_prime[j]
}
table_1_mr
table_1_mr_prime
table_1_mlm
table_1_replicate
```


